Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-12 paper.tex
Ditch the text graveyard

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark}, % VCFShark
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.

@@

Row-wise storage works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for all rows means retrieving and decompressing \emph{all} 
information from all rows.
The alternative to row-wise storage is \emph{columnar} storage:
instead of grouping together all the fields for a record,
we group together all the records for a given field.
Columnar storage has become standard [very quick summary of columnar
storage with some refs].
To our knowledge, the only existing method using columnar 
storage for genetic variation data is 
SeqArray~\citep{zheng2017seqarray,zheng2012high}.

@@

\section{Discussion}

Outline of a possible narrative by paragraph:
\begin{itemize}
\item When VCF was introduced as part of the work of the 1000 Genomes
project the software landscape was a real mess. It has entirely
succeeded in standardising the output of variant callers, and methods
have interoperated well for the past decade. As a method of interchange
following classical file-oriented paradigms it works well enough.
It is also a good archival format: when compressed with a well established
method like gzip and combined with the well-documented description of
the text, we can be sure that files will still be readable decades
from now.

\item VCF is holding back large-scale genomics, however, and in particular the
analysis of population scale genetic variation datasets like UKB and GeL.
Describe the current state of affairs. Variant datasets are chunked up
into reasonable sized bits (10s of gigabytes) so that they can be processed
in parallel. Doing anything with them requires interacting with workflow
managers. In practise, it is not possible to directly write programs to
work across multiple large-scale datasets: although the chunks of data
are given in VCF, orchestrating how the chunks are processed and
the results combined involves substantial work, involving the workflow
manager, and the details of how chromosomes are split into chunks.
Cite examples like the phasing pipelines for UKB from Brownings etc.

\item Distributed computing is the reality, and increasingly datasets
are being mandatorily stored in public clouds as part of
Trusted Research Environments. Cloud object stores are the reality
because it is much cheaper. POSIX file systems are emulated on top of
this. Benefits of breaking up data into manageable pieces are that
compute resources can be localised to the data. Compute resources are
most effectively obtained as short-lived, small VMs that operate
as part of a large cluster. Large vertically scaled ``pet'' servers
are much more expensive, and just not feasible for the current
scale of data. Lots of other architectural
benefits from the new reality of cloud-based storage.

\item Classical methods based around Unix pipelines and streaming
files between processes on a single machine are therefore a
significant mismatch to the reality of large-scale genetic variation
datasets. The majority of methods suggested to make working
with VCF data more efficient are based on these ideas: they all
create a single large file and regard distribution of work
in a cluster as extrinsic. Most output VCF as the result of queries,
which also assumes that the VCF data can be efficiently processed.
This is a reasonable assumption for thousands of samples, but not
for larger datasets.

\item The key issue here is that VCF is not a \emph{computational}
format. Contrast VCF with plink, which stores data on disk in a
form that can be computed on directly. Plink is limited in what
can be descibed though, does not compress, and is not designed
to distributed.

\item The large-scale analysis of genetic variation data does not
need a more efficient file format, which allows files to be downloaded
somewhat more quickly, and subsets extracted somewhat more efficiently
as VCF. It needs a standard storage platform that allows data to be
processed directly without conversion, subsetted efficiently across
both the variant and sample dimensions, with first-class support
for all the necessary additional information (GT values etc),
and that is suited to the realities of today's cloud-based
storage and processing facilities.

\item We suggest that the model used by Zarr is a good choice for this.
We have shown that chunks of the genotype matrix compressed with standard
methods are competitive with highly specialised methods for
genotype data. In reality most of the storage is for non-genotype
data anyway, where everyone is essentially doing the same thing.
Columnar storage allows getting just (e.g.) the variant positions
without reading everything. Another benefit is that by storing
fields separately in different cloud objects, they can have
different permissions. So, users can be given differential access
to read parts of the dataset, and this is managed directly
by the cloud storage. Give example.

\item A major strong point of Zarr is its simplicity. All it's doing
is compressing n-dimensional chunks of numerical data and storing
them using fixed address keys. This key-value approach means that storage
is very flexible and implementation is straightforward. There
are several implementations of the Zarr protocol, and while some of these
are rudimentary, they could be improved or replaced with relatively
little effort. Zarr has been successfully used to store petabytes
of data across different sciences --- genetic variation data is largely
arrays of numerical data, and not really that different ultimately.
Zarr is cloud native.
Zarr is an open specification, built on standard components (contrast
with Hail).

\item Zarr is not perfect. Discuss limitations around ragged arrays,
etc. Forthcoming improvements in zarr v3. The whole genotype matrix
paradigm is challenged by graph genomes as well. While this is
being worked out, there is still plenty of data to analyse as a
genotype matrix.

\item Vision. An ecosystem of tools that interoperate
based on an open, cloud-based format where distribution is naturally
managed by the stored chunks. Tools can be run on different datasets
with minimal tuning, not by writing a full layer of orchestration.
Tools read the data they need, not the entire dataset. Users read
directly from a single canonical dataset, and do not maintain
their own filtered copies. We suggest that the Zarr VCF approach
we have provided would be a good starting point for designing
such a compute platform.

\end{itemize}
[Random fragment to include somewhere]
Row-wise storage of the genotype matrix makes accessing 
sample haplotypes (i.e. the columns),
required for many
applications~\cite[e.g.][]{durbin2014efficient,kelleher2019inferring} 
[TODO add phasing, imputation]
inefficient. The simple approach of storing the genotype matrix
in rectangular chunks allows efficient access both by row \emph{and}
column. Similarly, these rectangular chunks capture patterns
both of similarity between samples and linkage disequilibrium~\cite{mcvean2019linkage} 
between sites, such that conventional compression schemes 
perform very well~\ref{fig-data-storage}.

Reel off Python data science ecosystem: Pandas~\citep{mckinney2010data} etc

@@

@@

@@

@@

\clearpage
\section*{Text graveyard}

Here's a bunch of text that contains bits that might be useful and readapted.

\subsection{Chunked storage and distributed computation}
How we store genetic data fundamentally affects how computation
can be structured, and ultimately the scalability of the calculations
we wish to perform. Classically, genetic variation data is stored via one
of two strategies. The first strategy, epitomised by
VCF~\citep{danecek2011variant} encodes all the data for a given
variant in a record, and then stores these records consecutively on file
(optionally compressed). This record-oriented approach maps well to the
powerful Unix pipelines abstraction and
has proven to be very successful.
The second classical strategy, epitomised
by the \texttt{plink} BED format~\citep{purcell2007plink}, is to
store genotype data in a structured binary format that can be
accessed directly. This memory-map approach tends to
be much faster to access, but leads to larger files and is more
limited in what can be represented. There are advantages and
disadvantages to both approaches, but one problem fundamentally
limits their utility for modern datasets: neither is well suited to
scaling-out across multiple machines.

[Probably need a paragraph here on what we've done classically
by splitting by variant. This was the dominant dimension, so that
worked fine. Now we need to also think about access-by-sample]

Given the scale of modern datasets, the ability to perform
calculations in parallel on multiple CPU cores
is now a basic prerequisite, and
running concurrently on multiple computers in a cluster
is increasingly becoming necessary. [For interactive analysis? People have
been doing concurrent computing with HPC clusters forever.]
Record-oriented variant storage such as VCF is not suitable
for  distributed computing because although it is relatively
straightforward to break workloads up by variant, we cannot efficiently
split by sample (see the section on Columnar Data below for details).
Storing data in a memory-mapped manner, where the genotypes
for each variant are written consecutive consecutively to a file,
suffers from the same problem. Although it is useful to conceptualise
the genotype matrix as two-dimensional, it is important to remember
that computer memory and file storage is ultimately one-dimensonal.
For example, plink's memory-mapped data layout linearises the 2D
genotype matrix by writing [TODO check] $n$ bytes representing
the genotypes for all samples at a variant position consecutively.
Because the genotypes for a given variant index can be found
in contiguous blocks at easily computable offsets from
this initial location in memory,
accessing the genotypes for a given variant is very efficient.
However, accessing the genotypes for a given sample at every
variant position is much less efficient because each of the $m$ bytes we
must read are $n$ bytes apart in the file.
In the distributed context, this often means
that each compute node  in a
cluster must effectively read the entire file in order to obtain the
data for a subset of samples. In practise, this means that calculations
are limited by I/0 bandwidth.

A simple and elegant solution to providing efficient access to such
large matrices across multiple dimensions is to store rectangular
\emph{chunks} rather than entire rows or columns. This is illustrated
in Fig~\ref{fig-data-storage}, where we show the basic approach
used by sgkit. By storing data in these regular chunks, we
can efficiently access data for both specific variants \emph{and}
samples.

Additionally, we can compress the individual chunks [more details].

This general strategy for storing multidimensional numerical data
has been in use for many years~\citep{folk2011overview}

This maps well to distributed computing because Zarr provides a
very flexible approach to storing the chunks. By using a native cloud store
for these chunks, compute nodes can directly access the chunks that
are needed, and not read irrelevant data.
[TODO flesh out and more stuff.]

\subsubsection{Array oriented computing}

NumPy~\citep{harris2020array} has been transformational, and provides
the basic framework for scientific computing in Python.

[ Things to cite: pandas~\citep{mckinney2010data},
BioNumpy~\citep{rand2022bionumpy}, OME-Zarr~\citep{moore2023ome}]

[NOTE: we may also want a section discussing the n-dimensional array model,
but I think this section is the right place to discuss that?]

\subsubsection{Columnar binary data}

Perhaps the most important factor limiting scalability in contemporary genomics
is the continued reliance on row-wise storage of data---most often
encoded as blockwise-compressed text.
[Explain why row-wise is bad without reference to specific genomics applications.]

The Variant Call Format (VCF) is the standard method for interchanging
genomics data~\citep{danecek2011variant}.
In VCF each row describes the observations for a set of samples
at a given position along the genome, and consists of fixed columns such as the
genome position as well as per-sample columns. [More info]
VCF was introduced as part of the 1000 Genomes project, and works reasonably
well at the scale of 1000s of samples. However, when we have hundreds of
thousands of samples it becomes extremely unwieldy
. For example, the
phased genotypes for 200K samples in UK Biobank produced by Browning [cite]
require X GB of space.
(Note that this file contains no extra INFO keys; in practice a great deal
more information is stored in the VCF leading to files that are orders of
magnitude larger, and requiring them to be split into chunks, further
increasing their unwieldiness.)
Running [a simple query that requires only looking at
one column] on this file like X required Y minutes.

[TODO be more concrete here; what is the thing we're summarising?]
This slow response time to a simple query is primarily caused by the row-wise
nature of VCF (and BCF), because in order to read a single piece of
information about a particular variant site we must read \emph{all}
the information
about that site. Then, if we wish to summarise that information over
all sites, then we must read the entire dataset.

A lesser but still significant cause of inefficiency is the use of
text to store the data. [More info]

These issues are of course well known, and there are numerous projects
that seek to alleviate them. [Discuss] They're all quite bespoke to genomics data
through.

In sgkit we use Zarr which although initially introduced to store
genomics data, has now been adopted widely in [X, Y and Z].
Zarr is [description] and has [advantages]

Also to discuss (not sure how these fit in to this specific sectio
on background technologies vs existing genomics infrastructure; can
sort out later):

\begin{itemize}
\item columnar (with small records) good for I/O elimination + sequential access.
Classical locality of reference stuff as well.
\item What does this mean with ND arrays? Slippery concept. Discuss.
\item Compression improvements from columnar.
\item Poor performance of text-based VCF has been noted many times,
and lots of prior work in this space e.g. \citep{kelleher2013processing}
\item Honourable mentions for bcftools~\citep{danecek2021twelve},
htslib~\citep{bonfield2021htslib}, CRAM~\citep{bonfield2014scramble,bonfield2022cram}
which uses a columnar approach
\end{itemize}

\subsection{Storing genetic variation data}

It is important to note that the example in Fig.~\ref{fig-data-storage}
is something of a best-base
scenario in terms genomic dataset storage, in that we are only
storing genotype data (per sample, per variant),
and, because it is error-free, is very
highly compressible. Noisier genotypes will not compress quite
so well, and storing less compressible data (such as genotype
quality values) results in a large increase in the overall
storage size. This is essentially unavoidable, but the Zarr-based
approach still has some significant advantages over VCF/BCF
[basically we can store separately and come up with distinct
compression strategies based on the data type. Dunno if we
want to get into this here. This is a complex topic with
a lot of existing work which we should review and reference,
and this is maybe not the place to do it.].

[SpVCF has some good text on high-entropy QC measures. We have to
reduce this entropy or we just can't store the data]

[Why are we reviewing all this? Well, we want to make the point
that people have been trying hard to address the problem of
how to deal with large-scale genomic data for quite a while,
and have come up with lots of specialised methods. There
are a few things to keep in mind. Our approach is to use
standard methods just on the "rectangles" of variant data
and it's already quite good. You could imagine writing specialised
compression filters for genotypes, if you wanted (maybe a
discussion item though). The key thing is the \emph{interface}.
They're all tying to the command line and to classical
sequential workflows. We have Python, and distributed compute. ]

Storage is the foundation on which any large-scale data-driven computation
rests: how we encode and organise data on persistent storage (the format)
fundamentally affects and limits the efficiencies of downstream calculations.
Different formats are optimised for different purposes, and there are
often many competing requirements when designing a data storage format.
% TODO is there some literature we could cite on this? I'm making up the
% term "computational format" here, presumably there is lots  prior thought
% on these ideas?
For our  purposes here, it is important to distinguish between \emph{archival}
and \emph{computational} formats. An archival format is optimised for
maximum interoperability and future-proofing, and uses the simplest and
most robust technologies to ensure the long-term usability of the data.
Reasonable compression levels are important because
of storage and transfer costs, but the efficiency of data access is a
secondary concern.
A computational format, on the other hand, is optimised
for calculations, for \emph{using} the data, today. A computational format
is designed with particular data access patterns in mind, ultimately
to make a broad range of queries and downstream computations as efficient as
possible.

% Note: a quick explanation of VCF for CS people coming in fresh
VCF~\citep{danecek2011variant} is a text-based format in which each line
describes observations
and metadata for a single variant (a position on the genome at which
variation occurs in the sample). The first eight tab-delimited
columns are mandatory, and subsequent columns describe the per-sample
genotype calls and quality control data.

[WIP: currently working on weaving together the stuff below into
some sort of narrative]

SeqArray~\citep{zheng2017seqarray} and the GDS format~\citep{zheng2012high}
provide efficient storage and programmatic access in R.

The approach taken by \toolname{plink}~\citep{purcell2007plink,chang2015second} is to
store data in an uncompressed binary format. This has some major
advantages.

BGEN \citep{band2018bgen} is optimised for storing large-scale
genotype array data and stores probabilities, and is used
in particular by UK Biobank~\citep{bycroft2018genome}
[Should consider the "hard-call" nature of most of these formats]

Many methods working with large scale genetic
data now use the PBWT or variants of
it~\citep[e.g.][]{li2016bgt,lefaive2021sparse,wertenbroek2022xsi}. PBWTs have
disadvantages, however: data must be phased, and compression level
is quite sensitive to the presence of errors [citation? people have
worked around this]

Several tools provide querying interfaces in their CLIs
(reminiscent of SQL) to provide easy access to particular samples, or
variants~\cite{li2016bgt,layer2016efficient}.

More recently XSI~\citep{wertenbroek2022xsi} is an file format that
uses a combination of strategies to achieve excellent compression
performance, and can support some calculations directly on the
compressed format. It provides a command line interface and a
C API.

The SAV format is an extension of BCF that takes advantage
of the sparseness of large scale genomic datasets~\citep{lefaive2021sparse}.
They provide a command line interface and a C++ API.
% TEXT from HSG chapter section on compressing data

% The simplest approach to compressing such data is to organise it in a
% `variant-centric' fashion, such that all observed genotypes for a given variant
% are stored in a contiguous block. Because most variants are rare, this will
% result in long runs of identical values that compress well using standard
% methods. This is the approach taken by the Variant Call
% Format~\citep{danecek2011variant}, and its more efficient binary encoding, BCF.
% The compression levels achieved by this straightforward approach are good, and
% quite competitive with more sophisticated methods. The disadvantage
% is that many queries require full decompression of the data, which can be
% prohibitively time-consuming. The SpeedGene data format and software
% library~\citep{qiao2012handling} chooses from one of three encoding methods for
% each SNP determined by the allele frequency; \cite{sambo2014compression} extend
% this idea by identifying blocks of SNPs in LD, and adding two additional
% potential encodings utilising this information.

% An alternative to this variant-centric approach is to store genotypes in a
% `sample-centric' fashion. Here, all genotypes for a particular sample are
% stored consecutively. While this breaks the simple repetition structure of data
% stored in variant-centric form, other methods can be employed to find
% compressible structure. For example, TGC~\citep{deorowicz2013genome} compresses
% sample genotypes using a custom Lempel-Ziv style approach on runs of identity
% among samples. This explicit use of LD structure results in excellent
% compression performance (for example, 32MB for all SNPs in 1000 Genomes
% chromosome 1); unfortunately, querying a dataset requires full decompression,
% making the format unsuitable for online analysis. The GQT
% toolkit~\citep{layer2016efficient} also uses the sample-centric organisation of
% genotype data, but takes a different approach to compression. Variants are
% sorted by allele frequency (resulting in longer runs of identical values within
% samples) and then compressed using an efficient bit-vector
% representation~\citep{wu2002compressing}. The resulting file-sizes are similar to
% compressed BCF, but many queries can be performed directly on the compressed
% representation and are therefore many times faster.

% The positional Burrows-Wheeler transform~\citep{durbin2014efficient}, or PBWT,
% is another sample-centric method. Building on the success of Burrows-Wheeler
% transform applications in
% genomics~\citep{langmead2009ultrafast,li2009fast,li2009soap2}, PBWT provides
% very good compression performance and efficient algorithms for finding
% haplotype matches. The algorithm builds a representation of the data based
% on sorted haplotype prefixes in linear time, and the LD structure of the data
% ensures that the sorting orders between adjacent sites change very little
% on average. The method has been successfully applied
% to phasing~\citep{loh2016reference}, detection of IBD
% segments~\citep{naseri2017ultra}, improving the performance of the Li and
% Stephens model~\citep{lunter2016fast}, and a general query engine for genotype
% data~\citep{li2016bgt}. Recent extensions include privacy preserving
% search~\citep{shimizu2016efficient} and generalisation to the setting of graph
% genomes~\citep{novak2017graph}.

\subsection{Distributed computation}
Distributed computation has been the reality in the analysis
of genetic variation for many years,
where data is most often provided as per-chromosome VCFs.
This chunking provides a fairly straightforward way to parallelise
computations, using HPC schedulers to split work per-file.
When more fine-grained parallelism is required, users must
provide exact genome coordinates of the genome regions
as part of their submission scripts. Combining results
is a manual operation.
Workflow engines such as
Snakemake~\cite{koster2012snakemake,molder2021sustainable}
and [TODO mention WDL, Cromwell, etc?] make such analyses
much easier, but there is still a great deal of room for
mistakes.
Per-chromosome VCFs are not practical for biobank scale
projects, however, and variant call data tends to be split
into many (still large) files.
For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
hundreds of files per chromosome.
% This is horrible writing, but the point is important.
While these files provide a natural
unit for distributing work, the details of how they
are split differ and essentially there is
no interoperability among large-scale datasets because the VCFs are
so cumbersome that a separate orchestration layer is required to
access the data in parallel.

Despite this long history of de-facto distributed computation
and the clear need for better methods of accessing genetic
variation data in a distributed fashion, there is almost
no scientific literature on the topic (in stark contrast to the
study of methods to compress a VCF into a single file,
as discussed in section XXX).

[This is a really rough outline, also not sure how exactly
we sequence the ideas here. Just getting paragraph fragments
on paper. We do need to explain something of the hadoop ecosystem
to readers who are not familiar, though.]
Distributed analytics for large-scale data is a mature
techology, with many different approaches. Hadoop~\citep{white2012hadoop}
is the original method, and remains the basis of the
dominant ecosystem. HDFS. Spark [CITATION?] lets us
do more complex analytics. Parquet [CITATION?]

Although the Hadoop ecosystem has been hugely successful
in many domains, it has had a limited impact in genomics.
[List some things that have been tried for e.g variant calling
, like ADAM~\citep{nothaft2015rethinking}].
[Briefly mention two approaches using Hadoop clusters
~\citep{boufea2017managing,fan2020variant}, to not very
exiting results].

[Hadoop/Spark/Parquet are fundamentally unsuited to genetic
variation data [WHY]?]

Discuss Hail. It's awesome, has enabled current
generation of mega-scale projects. Drawbacks are the
architecture has emphasised performance etc over adaptability
and simplicity. Uses bits of Spark for X, but custom YZ .
File format is undocumented, etc.

Discuss why the sgkit architecture is suited to distributed
computation. Discuss Zarr's use in cloud stores, and existing
deployments. Discuss Dask.
We do not attempt a head-to-head performance comparison with
Hail here, but defer to Liangde's thesis~\citep{li2022efficient}
with brief summary.
Forward ref to following sections on applications to demonstate
sgkit in a distributed context.

>>>>

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark}, % VCFShark
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.

@@

Row-wise storage works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for all rows means retrieving and decompressing \emph{all} 
information from all rows.
The alternative to row-wise storage is \emph{columnar} storage:
instead of grouping together all the fields for a record,
we group together all the records for a given field.
Columnar storage has become standard [very quick summary of columnar
storage with some refs].
Although columnar methods have been successfully used in alignment 
% TODO check
storage~\citep{bonfield2014scramble,bonfield2022cram},
the only existing method using columnar 
storage for genetic variation data is 
SeqArray~\citep{zheng2017seqarray,zheng2012high} to our knowledge.

@@

\section{Discussion}
The introduction of VCF as part of the work of the 1000 Genomes
project~\cite{10002015global} has transformed a fragmented

% A computational format, on the other hand, is optimised
% for calculations, for \emph{using} the data, today. A computational format
% is designed with particular data access patterns in mind, ultimately
% to make a broad range of queries and downstream computations as efficient as
% possible.

% Outline of a possible narrative by paragraph:
% \begin{itemize}
% \item When VCF was introduced as part of the work of the 1000 Genomes
% project the software landscape was a real mess. It has entirely
% succeeded in standardising the output of variant callers, and methods
% have interoperated well for the past decade. As a method of interchange
% following classical file-oriented paradigms it works well enough.
% It is also a good archival format: when compressed with a well established
% method like gzip and combined with the well-documented description of
% the text, we can be sure that files will still be readable decades
% from now.

% \item VCF is holding back large-scale genomics, however, and in particular the
% analysis of population scale genetic variation datasets like UKB and GeL.
% Describe the current state of affairs. Variant datasets are chunked up
% into reasonable sized bits (10s of gigabytes) so that they can be processed
% in parallel. Doing anything with them requires interacting with workflow
% managers. In practise, it is not possible to directly write programs to
% work across multiple large-scale datasets: although the chunks of data
% are given in VCF, orchestrating how the chunks are processed and
% the results combined involves substantial work, involving the workflow
% manager, and the details of how chromosomes are split into chunks.
% Cite examples like the phasing pipelines for UKB from Brownings etc.

% \item Distributed computing is the reality, and increasingly datasets
% are being mandatorily stored in public clouds as part of
% Trusted Research Environments. Cloud object stores are the reality
% because it is much cheaper. POSIX file systems are emulated on top of
% this. Benefits of breaking up data into manageable pieces are that
% compute resources can be localised to the data. Compute resources are
% most effectively obtained as short-lived, small VMs that operate
% as part of a large cluster. Large vertically scaled ``pet'' servers
% are much more expensive, and just not feasible for the current
% scale of data. Lots of other architectural
% benefits from the new reality of cloud-based storage.

% \item Classical methods based around Unix pipelines and streaming
% files between processes on a single machine are therefore a
% significant mismatch to the reality of large-scale genetic variation
% datasets. The majority of methods suggested to make working
% with VCF data more efficient are based on these ideas: they all
% create a single large file and regard distribution of work
% in a cluster as extrinsic. Most output VCF as the result of queries,
% which also assumes that the VCF data can be efficiently processed.
% This is a reasonable assumption for thousands of samples, but not
% for larger datasets.

% \item The key issue here is that VCF is not a \emph{computational}
% format. Contrast VCF with plink, which stores data on disk in a
% form that can be computed on directly. Plink is limited in what
% can be descibed though, does not compress, and is not designed
% to distributed.

% \item The large-scale analysis of genetic variation data does not
% need a more efficient file format, which allows files to be downloaded
% somewhat more quickly, and subsets extracted somewhat more efficiently
% as VCF. It needs a standard storage platform that allows data to be
% processed directly without conversion, subsetted efficiently across
% both the variant and sample dimensions, with first-class support
% for all the necessary additional information (GT values etc),
% and that is suited to the realities of today's cloud-based
% storage and processing facilities.

% \item We suggest that the model used by Zarr is a good choice for this.
% We have shown that chunks of the genotype matrix compressed with standard
% methods are competitive with highly specialised methods for
% genotype data. In reality most of the storage is for non-genotype
% data anyway, where everyone is essentially doing the same thing.
% Columnar storage allows getting just (e.g.) the variant positions
% without reading everything. Another benefit is that by storing
% fields separately in different cloud objects, they can have
% different permissions. So, users can be given differential access
% to read parts of the dataset, and this is managed directly
% by the cloud storage. Give example.

% \item A major strong point of Zarr is its simplicity. All it's doing
% is compressing n-dimensional chunks of numerical data and storing
% them using fixed address keys. This key-value approach means that storage
% is very flexible and implementation is straightforward. There
% are several implementations of the Zarr protocol, and while some of these
% are rudimentary, they could be improved or replaced with relatively
% little effort. Zarr has been successfully used to store petabytes
% of data across different sciences --- genetic variation data is largely
% arrays of numerical data, and not really that different ultimately.
% Zarr is cloud native.
% Zarr is an open specification, built on standard components (contrast
% with Hail).

% \item Zarr is not perfect. Discuss limitations around ragged arrays,
% etc. Forthcoming improvements in zarr v3. The whole genotype matrix
% paradigm is challenged by graph genomes as well. While this is
% being worked out, there is still plenty of data to analyse as a
% genotype matrix.

% \item Vision. An ecosystem of tools that interoperate
% based on an open, cloud-based format where distribution is naturally
% managed by the stored chunks. Tools can be run on different datasets
% with minimal tuning, not by writing a full layer of orchestration.
% Tools read the data they need, not the entire dataset. Users read
% directly from a single canonical dataset, and do not maintain
% their own filtered copies. We suggest that the Zarr VCF approach
% we have provided would be a good starting point for designing
% such a compute platform.

% \end{itemize}

[Random fragment to include somewhere]
Row-wise storage of the genotype matrix makes accessing 
sample haplotypes (i.e. the columns),
required for many
applications~\cite[e.g.][]{durbin2014efficient,kelleher2019inferring} 
[TODO add phasing, imputation]
inefficient. The simple approach of storing the genotype matrix
in rectangular chunks allows efficient access both by row \emph{and}
column. Similarly, these rectangular chunks capture patterns
both of similarity between samples and linkage disequilibrium~\cite{mcvean2019linkage} 
between sites, such that conventional compression schemes 
perform very well~\ref{fig-data-storage}.

Reel off Python data science ecosystem: Pandas~\citep{mckinney2010data} etc
[ Things to cite: pandas~\citep{mckinney2010data},
BioNumpy~\citep{rand2022bionumpy}, OME-Zarr~\citep{moore2023ome}]

Although the Hadoop ecosystem has been hugely successful
in many domains, it has had a limited impact in genomics.
[List some things that have been tried for e.g variant calling
, like ADAM~\citep{nothaft2015rethinking}].
[Briefly mention two approaches using Hadoop clusters
~\citep{boufea2017managing,fan2020variant}, to not very
exiting results].
Distributed analytics for large-scale data is a mature
techology, with many different approaches. Hadoop~\citep{white2012hadoop}
is the original method, and remains the basis of the
dominant ecosystem. HDFS. Spark [CITATION?] lets us
do more complex analytics. Parquet [CITATION?]
Distributed computation has been the reality in the analysis
of genetic variation for many years,
where data is most often provided as per-chromosome VCFs.
This chunking provides a fairly straightforward way to parallelise
computations, using HPC schedulers to split work per-file.
When more fine-grained parallelism is required, users must
provide exact genome coordinates of the genome regions
as part of their submission scripts. Combining results
is a manual operation.
Workflow engines such as
Snakemake~\cite{koster2012snakemake,molder2021sustainable}
and [TODO mention WDL, Cromwell, etc?] make such analyses
much easier, but there is still a great deal of room for
mistakes.
Per-chromosome VCFs are not practical for biobank scale
projects, however, and variant call data tends to be split
into many (still large) files.
For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
hundreds of files per chromosome.
% This is horrible writing, but the point is important.
While these files provide a natural
unit for distributing work, the details of how they
are split differ and essentially there is
no interoperability among large-scale datasets because the VCFs are
so cumbersome that a separate orchestration layer is required to
access the data in parallel.

@@

% Zarr was originally developed as a simpler version of 
% HDF5~\citep{folk2011overview}

@@

The vcfzarr specification is general and can be mapped to 
file formats such as \toolname{plink}~\citep{purcell2007plink,chang2015second}
and BGEN~\citep{band2018bgen} with some minor extensions.

@@

Fields such as PL are currently problematic when there are large 
numbers of alleles. However, we plan on implementing the 
local alleles strategy~\citep{poterba2024scalable,danecek2021twelve}
which will likely become part of the VCF specification.

@@
